---
title: "R for Data Science: Data transformation"
output: html_notebook
---

```{r}
library(nycflights13)
library(tidyverse)
```

### 5.2.4 Exercises

#### 1. Find all flights that
```{r}
str(flights)
```

##### 1.1. Had an arrival delay of two or more hours
```{r}
filter(flights, arr_delay >= 120)
```

##### 1.2. Flew to Houston (IAH or HOU)
```{r}
filter(flights, dest == "IAH" | dest == "HOU")
```

##### 1.3. Were operated by United, American, or Delta
```{r}
# for airline names see http://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=236 and download carroer CSV
# "AA","American Airlines Inc. (1960 - )"
# "UA","United Air Lines Inc. (1960 - )"
# "DL","Delta Air Lines Inc. (1960 - )"
filter(flights, carrier %in% c("AA","UA","DL"))
```

##### 1.4. Departed in summer (July, August, and September)
```{r}
filter(flights, month %in% c(7,8,9))
```

##### 1.5. Arrived more than two hours late, but didn’t leave late
```{r}
filter(flights, arr_delay >= 120 & dep_delay == 0)
```

##### 1.6. Were delayed by at least an hour, but made up over 30 minutes in flight
```{r}
filter(flights, dep_delay >= 60 & dep_delay - arr_delay > 30)
```


##### 1.7. Departed between midnight and 6am (inclusive)
```{r}
filter(flights, dep_time %/% 100 %in% c(0:5) | dep_time == 600)
```


#### 2. Another useful dplyr filtering helper is between(). What does it do? Can you use it to simplify the code needed to answer the previous challenges?

```{r}
# This is a shortcut for x >= left & x <= right
# function (x, left, right) 
# {
#  .Call("dplyr_between", PACKAGE = "dplyr", x, left, right)
# }
filter(flights, between(month,7,9))
filter(flights, between(dep_time %/% 100, 0, 5) | dep_time == 600)
```


#### 3. How many flights have a missing dep_time? What other variables are missing? What might these rows represent?

The dep_time is missing as well as air_time, dep_delay, arr_time and arr_delay. These look like cancelled flights.
```{r}
filter(flights, is.na(dep_time))
```

#### 4. Why is NA ^ 0 not missing? Why is NA | TRUE not missing? Why is FALSE & NA not missing? Can you figure out the general rule? (NA * 0 is a tricky counterexample!)

```{r}
# Anything power of zero is one. The evaluation is "dep_time == 1"
filter(flights, dep_time == NA ^ 0 )
```
```{r}
# Anything or true is always true. Thus returning all entries.
filter(flights, dep_time == NA | TRUE )
```
```{r}
# False and anything is always false. Thus returning nothing.
filter(flights, dep_time == FALSE & NA )
```

```{r}
# NA within a calculation always results to NA. The dep_time cannot be compare to an unknown value. Thus resulting nothing.
filter(flights, dep_time == NA * 0)
```

### 5.3.1 Exercises

#### 1. How could you use arrange() to sort all missing values to the start? (Hint: use is.na()).
```{r}
arrange(flights, desc(is.na(dep_time)))
# Use order(na.last = FALSE) directly to order all values?
```

#### 2. Sort flights to find the most delayed flights. Find the flights that left earliest.
```{r}
arrange(flights, desc(dep_delay))
arrange(flights, dep_delay)
```


#### 3. Sort flights to find the fastest flights.
```{r}
arrange(flights, desc(distance %/% air_time))
```


#### 4. Which flights travelled the longest? Which travelled the shortest?
```{r}
# OK, its easier to look at the air-time. Otherwise we have to handle the time ints.
arrange(flights, desc(air_time))
arrange(flights, air_time)
```


### 5.4.1 Exercises

#### 1. Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.
```{r}
select(flights, dep_time, dep_delay, arr_time, arr_delay)
select(flights, dep_time, arr_time, ends_with("delay"))
select(flights, dep_time, arr_time, contains("delay"))
# exclude with -() 
# not even starting with regex...
```


#### 2. What happens if you include the name of a variable multiple times in a select() call?
```{r}
# Nothing special happens.
select(flights, dep_time, dep_time)
```


#### 3. What does the one_of() function do? Why might it be helpful in conjunction with this vector?

```{r}
# No idea yet
# one_of(..., vars = current_vars())
# everything(vars = current_vars())
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
select(flights, one_of(vars))
```


#### 4. Does the result of running the following code surprise you? How do the select helpers deal with case by default? How can you change that default?

```{r}
# contains(match, ignore.case = TRUE, vars = current_vars())
select(flights, contains("TIME"))
```

